Collaborative Infrastructure for Test-Driven\\Scientific Model Validation

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: Final draft

\usepackage{enumitem}
\setenumerate{noitemsep,topsep=2pt,parsep=0pt,partopsep=0pt}
\newcommand{\subparagraph}{}
\usepackage[compact]{titlesec}
\titlespacing{\section}{0pt}{*0}{*0}
\titlespacing{\subsection}{0pt}{*0}{*0}
\titlespacing{\subsubsection}{0pt}{*0}{*0}

@@

We argue that a similar test-driven methodology would be valuable to scientific communities as they seek to  validate increasingly complex models against growing repositories of empirical data. Scientific communities differ from software communities in several key ways, however. In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation and outline how SciUnit, supported by new and existing collaborative infrastructure, could be  integrated into the modern scientific workflow.
\end{abstract}
\category{D.2.5}{Software Engineering}{Testing and Debugging}%[complexity measures, performance measures]
\terms{Documentation; measurement; standardization}
\keywords{unit testing; model validation; cyberinfrastructure}

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
%\IEEEpeerreviewmaketitle

\section{Introduction}
\begin{figure*}[t]
\vspace{-30px}
\centering
\includegraphics[scale=0.45]{diagram1.pdf}
\vspace{-50px}
\caption{\small Tests are derived from data and models are derived from scientific theories, either directly or through existing data and model repositories. The score table summarizes the performance of a collection of models against a suite of tests. Tests and models are initialized, executed and visualized using \texttt{sciunit} in an IPython notebook stored inside a \emph{suite repository} (e.g. \texttt{saturnsuite}) hosted on a social coding website (e.g. Github). SciDash  discovers and organizes  suite repositories and provides interactive views of the notebooks they contain. Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed (e.g. \texttt{cosmounit}).}  
\label{fig:sciunit_overview}
\vspace{-10pt}
\end{figure*}

@@

Before explaining the details of this example, we point out that SciUnit is implemented in Python. Python is one of the most widely used languages in science today \cite{sanner1999python}, and is increasingly regarded as the \textit{de facto} language of open source scientific computing. It is easy to read and has a simple object system that we use to specify the interface between tests and models (challenge 2, see below). It supports calling into other popular languages, including R, MATLAB, C and Java, often more cleanly than they support calling into Python (challenge 3). The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we will use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{ipynb}. Together, this makes writing wrappers around tests and models written in other languages relatively simple. We anticipate making it nearly automatic as future work.
\begin{figure}[t]
\small
%class PositionTest(sciunit.Test):
%    """Tests a planetary position model based on positions observed on night n given the positions in the n-1 previous nights.
%    Metric: Standard p-value.
%    Parameters: 
%      obs_histories : list[list[position]]
%      obs_positions: list[position]"""
%	def __init__(self, obs_histories, obs_positions):
%		self.obs_histories = obs_histories
%		self.obs_positions = obs_positions
%
%	required_capabilities = [PredictsPlanetaryPosition]
%
%	def _judge(self, model):
%		predictions = []
%		for obs_history in self.obs_histories:
%			predictions.append(model.predict_next_pos(obs_history))
%		p = pooled_p_val(predictions, self.obs_positions)
%		return sciunit.PValue(p, related_data={
%			'obs_histories': obs_histories,
%			'obs_positions': obs_positions,
%			'predictions': predictions
%		})
\begin{python}
class PositionTest(sciunit.Test):
  """Tests a planetary position model based on positions observed on night n given the positions in the n-1 previous nights.
  Observation: {
    'histories':list[list[Position]], 
    'positions':list[Position]]
  }"""
  required_capabilities = [PredictsPlanetaryPosition]
  def generate_prediction(self, model):
    return [model.predict_next_pos(obs_history)
      for obs_history 
      in self.observation['histories']]

@@

Social coding tools like Github allow each scientist to work on their own fork of a repository where they can examine the consequences of their own contributions on  the tests and models that have previously been developed, facilitating both model and data sharing, and the organic adoption of community standards. For example, an experimentalist who has gathered new data can instantiate a new test or refine an existing one  with more accurate observations. As long as the interface between the test and the model remains the same, the scores can be recomputed entirely automatically, requiring no coordination with the original model developers (challenge 2). Similarly, when a modeler develops a new model, she can immediately  validate it against all the tests it is capable of taking. If a statistician develops a better goodness-of-fit metric, it too can be turned into a collection of tests that use existing capabilities, and so the table can be regenerated automatically, as discussed above. If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).

Once satisfied, these scientists might submit a \emph{pull request} to the central repository (either separately or in conjunction with the submission of a conventional paper). Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before committing it to the ``academic record''. A model that performs well across tests in a collaboratively developed suite could more easily claim (e.g. to reviewers) to be a coherent, valid model of, for example, Saturn's motion.

@@

To help help scientists discover these repositories and agree on which ones are central, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}). It is essentially a wiki pointing scientists to the $X$\verbx{unit}and $X$\verbx{suite} repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well). SciDash also supports loading IPython notebooks containing generated tables stored in these repositories into the browser directly so that score tables can be explored without going to the trouble of cloning the repository locally. To modify the suite (e.g. by adding the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of such a view. We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git and installation of Python and the packages required by the repository are left behind the scenes (challenge 4). SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available.

TODO: How to get adoption? Run contests -- dash double meaning. Also ask people directly (worked for Scholarpedia...) We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields. Typically such competitions use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible.

@@

\section{Discussion}\label{discussion}
The design just described has been purposefully left as simple as possible, in pursuit of challenge 4. Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out. TODO: SciUnit is essentially an embedded DSL that captures precisely the standard workflow of model validation. SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.

@@

While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in a project called NeuronUnit (http://github.com/scidash/neuronunit). We have used this as part of a collaboration with a large-scale neuroinformatics project (http://www.opensourcebrain.org), whose developers are enthusiastic about a practical pipeline for testing the dozens of models they currently maintain. TODO: More SE-research-friendly future work discussion Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies. Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing. The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/. SciDash is under active development, and is currently capable of basic forms of the operations described in Sec. 3.

 %In particular, we consider modeling competitions in science as a form of test-driven validation. For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes, and 
%examples from other subfields of biology abound (e.g. \url{http://www.the-dream-project.org}).  These competitions typically use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible. In related work that

@@

\section*{Acknowledgements}
We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their helpful discussions.  
The work was supported in part by NIMH grant R01MH081905. 
The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.

% conference papers do not normally have an appendix

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{abbrv}
\bibliography{../research,references,references_crook,urls,references_ja}
%\bibliography{references,references_crook,urls,references_ja}

>>>>

\usepackage{enumitem}
\setenumerate{noitemsep,topsep=2pt,parsep=0pt,partopsep=0pt}
\newcommand{\subparagraph}{}
\usepackage[compact]{titlesec}
\titlespacing{\section}{0pt}{*0.2}{*0}
%\titleformat{\section}
%  {\large\bfseries\uppercase}{\thesection}{1em}{}
\titlespacing{\subsection}{0pt}{*0}{*0}
\titlespacing{\subsubsection}{0pt}{*0}{*0}

@@

We argue that a similar test-driven methodology would be valuable to scientific communities as they seek to  validate increasingly complex models against growing repositories of empirical data. Scientific communities differ from software communities in several key ways, however. In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation and outline how SciUnit, supported by new and existing collaborative infrastructure, could be  integrated into the modern scientific workflow.
\end{abstract}
\category{D.2.5}{Software Engineering}{Testing and Debugging}%[complexity measures, performance measures]
\terms{Documentation, Measurement, Standardization}
\keywords{unit testing, model validation, cyberinfrastructure}

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
%\IEEEpeerreviewmaketitle
\titleformat{\section}
 {\secfnt}{\thesection}{1em}{\secfnt\MakeUppercase}
 \vspace{10pt}
\section{Introduction}
\begin{figure*}[t]
\vspace{-30px}
\centering
\includegraphics[scale=0.45]{diagram1.pdf}
\vspace{-50px}
\caption{\small Tests are derived from data and models are derived from scientific theories, either directly or through existing data and model repositories. The score table summarizes the performance of a collection of models against a suite of tests. Tests and models are initialized, executed and visualized using \texttt{sciunit} in an IPython notebook stored inside a \emph{suite repository} (e.g. \texttt{saturnsuite}) hosted on a social coding website (e.g. Github). SciDash  discovers and organizes  suite repositories and provides interactive views of the notebooks they contain. Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed (e.g. \texttt{cosmounit}).}  
\label{fig:sciunit_overview}
\vspace{-10pt}
\end{figure*}

@@

Before explaining the details of this example, we point out that SciUnit is implemented in Python. Python is one of the most widely used languages in science today \cite{sanner1999python}, and is increasingly regarded as the \textit{de facto} language of open source scientific computing. It is easy to read and has a simple object system that we use to specify the interface between tests and models (challenge 2, see below). It supports calling into other popular languages, including R, MATLAB, C and Java, often more cleanly than they support calling into Python (challenge 3). The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we will use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{perez2007ipython}. Together, this makes writing wrappers around tests and models written in other languages relatively simple. We anticipate making it nearly automatic as future work.
\begin{figure}[t]
\small
%class PositionTest(sciunit.Test):
%    """Tests a planetary position model based on positions observed on night n given the positions in the n-1 previous nights.
%    Metric: Standard p-value.
%    Parameters: 
%      obs_histories : list[list[position]]
%      obs_positions: list[position]"""
%	def __init__(self, obs_histories, obs_positions):
%		self.obs_histories = obs_histories
%		self.obs_positions = obs_positions
%
%	required_capabilities = [PredictsPlanetaryPosition]
%
%	def _judge(self, model):
%		predictions = []
%		for obs_history in self.obs_histories:
%			predictions.append(model.predict_next_pos(obs_history))
%		p = pooled_p_val(predictions, self.obs_positions)
%		return sciunit.PValue(p, related_data={
%			'obs_histories': obs_histories,
%			'obs_positions': obs_positions,
%			'predictions': predictions
%		})
\begin{python}
class PositionTest(sciunit.Test):
  """Tests a planetary position model based on positions observed on night n given the positions in the n-1 previous nights.
  Observation: {
    'histories':list[list[Position]], 
    'positions':list[Position]]
  }"""
  required_capabilities = [PredictsPlanetaryPosition]
  def generate_prediction(self, model):
    return [model.predict_next_pos(obs_history)
      for obs_history 
      in self.observation['histories']]

@@

Social coding tools like Github allow each scientist to work on their own fork of a repository like \verbx{saturnunit} where they can examine the consequences of their own contributions on  the tests and models that have previously been developed. For example, an experimentalist who has gathered new data can instantiate a new test or refine an existing one  with more accurate observations. As long as the interface between the test and the model remains the same, the scores for prior models can be recomputed entirely automatically, requiring no coordination with model developers (challenge 4). When a modeler develops a new model, she can   validate it against all the tests it is capable of taking. If a statistician develops a better goodness-of-fit metric, it too can existing capabilities, and so a comprehensive score table can be generated easily, as discussed above. If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).

Once satisfied with their contributions to the repository, these scientists might submit a \emph{pull request} to the central repository (either separately or together with the submission of a conventional paper). Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before committing it to the ``academic record''. A model that performs well across tests in a collaboratively developed suite could more directly and believably claim (e.g. to reviewers) to be a coherent, valid model of, for example, Saturn's motion.

@@

To help scientists discover and coordinate control over these repositories, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}). It is essentially a wiki pointing scientists to the common and suite repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well). SciDash also supports loading IPython notebooks containing score tables stored in these repositories into the browser directly, avoiding the need to clone the repository locally. To modify the suite (e.g. to add the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of this view. We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git and installation of Python and the packages required by the repository are left behind the scenes (challenge 4). SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available, fostering collaboration.

As with many pieces of cyberinfrastructure, early adopters may benefit less than late adopters due to network effects. To incentivize adoption, we look to an increasingly common scientific practice: modeling competitions. Many research communities use these to periodically determine the state of the art, resulting in a form of review paper. Funding sources often include small monetary incentives.
%For example, the quantitative single neuron modeling competition operated over the course of two years  \cite{qsnmc}. 
   SciDash could be used to run these kinds of competitions on a continuous basis atop less \textit{ad hoc} infrastructure (and thus require less effort and central planning from organizers) and we are actively organizing such competitions.

@@

\vspace{5px}
\section{Discussion \& Future Directions}\label{discussion}
Instead of attempting to use an existing general-purpose testing infrastructure, we have designed what is essentially a domain-specific language that captures the scientific model validation process in more familiar terms (challenge 6). SciUnit can be used by individual scientists to organize their workflows and also within a loosely collaborative workflow, mediated by a social coding tool like Github and an easy-to-use portal, SciDash, based on the IPython system \cite{perez2007ipython}.

@@

While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with a large-scale informatics project, OpenSourceBrain, whose developers are enthusiastic about developing a practical pipeline for precisely characterizing the scope and validity of the dozens of models they currently maintain (case study in preparation). The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate quantitative models of, for example, software cost or component reusability.

This paper identifies potential  synergies between software development and model validation practices, and provides a proof of concept to argue that a test-driven development methodology could help improve scientific workflows. The modern scientific model validation process has not yet been extensively studied by  the research community. Precisely characterizing how scientists today attempt to answer the questions in the introduction, and evaluating the effectiveness of tools like the ones we propose, are likely fruitful avenues for future empirical research.

The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/. SciDash is under active development as of this writing.
\vspace{5px}
 %In particular, we consider modeling competitions in science as a form of test-driven validation. For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes, and 
%examples from other subfields of biology abound (e.g. \url{http://www.the-dream-project.org}).  These competitions typically use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible. In related work that

@@

\section{ACKNOWLEDGEMENTS}
We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions.  
This work was supported in part by NIMH grant R01MH081905. 
The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.
\vspace{5px}
% conference papers do not normally have an appendix

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\renewcommand\bibsection{\subsection{\refname}}
\renewcommand\refname{New Title}
\bibliographystyle{abbrv}
\bibliography{../research,references,references_crook,urls,references_ja}
%\bibliography{references,references_crook,urls,references_ja}
